Chemical Tagging Using Minimum Message Length. I.\\ Gaussian Mixture Modelling

### 2017-04-20 articles/chemical-tagging-gmm/ms.tex
enough for now

In this \article\ we will introduce a simple probabilistic approach to
chemical tagging, where the data are modelled as a mixture of multivariate
Gaussian distributions and the number of mixtures is unknown (Section 
\ref{sec:the-model}). 
In Section \ref{sec:mml} we introduce the principle of minimum message length
(MML) and formulate our objective function.
Although our model is simplistic in the sense that we will only address some
aspects of chemical tagging, in Section \ref{sec:experiments} our experiments
show that our MML-based approach is superior to all other methods considered 
for chemical tagging.
In Section \ref{sec:discussion} we describe how the MML principle can be used
to construct a fully consistent probabilistic approach to chemical tagging,
and simultaneously address the challenges outlined in Section 
\ref{sec:the-problem}.

@@

Background to MML, etc.

We want to minimize the message length.

Link to Bayesian evidence, other things, etc.

@@

\bibitem[Schafer (1997)]{Schafer_1997} Schafer, J.~L., 1997, Analysis of Incomplete Multivariate Data (CRC Press) % 10.1002/(SICI)1097-0258(20000415)19:7<1006::AID-SIM384>3.0.CO;2-T

>>>>

In this \article\ we will introduce a simple probabilistic approach to
chemical tagging, where the data are modelled as a mixture of multivariate
Gaussian distributions and the number of mixtures is unknown.
In Section \ref{sec:the-model} we describe our model and outline our objective
function.
Although our model is simplistic in the sense that we will only address some
aspects of chemical tagging, in Section \ref{sec:experiments} our experiments
show that our approach is superior to all other methods considered 
for chemical tagging.
In Section \ref{sec:discussion} we describe how the principles introduced here
can be used to construct a fully consistent probabilistic approach to chemical
tagging, and simultaneously address the challenges outlined in Section 
\ref{sec:the-problem}.

@@

The Minimum Message Length \citep[MML;][]{Wallace_1968} principle is a formal
description of Occam's razor that is based on information theory.
The MML principle states that the best explanation of the data is the model
that describes the data using the smallest amount of information (shortest
message).
Information is described with the same description as \citet{Shannon_1948},
who showed that that given an event $E$ with probability $P(E)$, the length of 
the shortest lossless code $I(E)$ to represent that event requires 
$I(E) = -\log_{2}P(E)$ bits of information\footnote{
  Unless specified -- as it is in this instance -- throughout this article we 
  will refer to the information $I$ or entropy in natural units (nats). 
  Recall that 1 bit $= \log{2}$ nats.
  }.
\citet{Wallace_1968} linked this result with Bayes' theorem,

\begin{equation}
  P(H|D) = \frac{P(H)\cdot{}P(D|H)}{P(D)}
\end{equation}

\noindent{}which states that the joint probability of the data $D$ and 
hypothesis $H$, $P(H, D)$, is given by

\begin{equation}
  P(H, D) = P(H) \cdot P(D|H) = P(D) \cdot P(H|D)
\end{equation}

\noindent{}where $P(H)$ is the prior probability of the hypothesis $H$,
$P(D|H)$ is the likelihood, $P(D)$ is the prior probability of the data $D$,
and $P(H|D)$ is the posterior probability of the hypothesis given the data.
Specifically, \citet{Wallace_1968} derived the relationship between
conditional probabilities in terms of optimal message lengths,

\begin{equation}
  I(H, D) = I(H) + I(D|H) = I(D) + I(H|D) \quad .
  \label{eq:mml-wallace}
\end{equation}

Therefore the total cost of a message is the sum of the message length of the
hypothesis $H$, which takes $I(H)$ bits, and the data $D$ given the hypothesis
$H$, $I(D|H)$. 
Considering a simple problem of model selection provides good intuition for
this two-part message approach. 
Consider a simple model $H_0$ where the first part of the message $I(H_0)$ is 
short, and which provides a poor fit to the data such that $I(D|H_0)$ is long.
An alternative model $H_1$ is more complex, requiring a longer $I(H_1)$, but
the increased model flexibility provides a better fit to the data, making
$I(D|H_1)$ short.
For both parameter estimation \emph{and} model selection, we want to chose
$H$ (of $H_0$, $H_1$, etc) and the model parameters that minimizes the
\emph{total} message length.

Calculating the length of the message is a non-trivial task, especially if the
model is even reasonably complex.
This makes the MML principle intractable (or uncomputable) in most cases, and
forces us to make approximations when calculating the message length.
\citet{Wallace_Freeman_1987} introduce a generalised scheme to estimate a
vector of parameters $\vectheta$ of any distribution given some data, where
the estimates $\vectheta$ that minimise the message length expression,

\begin{equation}
  I(\vectheta,\data) 
    = \frac{p}{2}\log{q_p} - \log{\left(\frac{h(\vectheta)}{\sqrt{|\mathcal{F}(\vectheta)|}}\right)}
      - \mathcal{L}\left(\data|\vectheta\right) + \frac{p}{2}
\end{equation}

\noindent{}is the MML estimate.
Here $h(\vectheta)$ is the prior on the hypothesis, $p$ is the number of free
parameters in the model, and $|\mathcal{F}\left(\vectheta\right)|$ is the
determinant of the \emph{expected} Fisher information matrix (the second-order
partial derivatives of the negative log-likelihood function $-\mathcal{L}(\data|\vectheta)$.

The message length for a model includes the length (in bits of entropy)
required to transmit the model parameters, and the length required to transmit
the residuals of the data given the model.
There is a wealth of literature on the successes (and derivatives) of MML
in the fields of medicine, statistics, etc.

- Deep grounds in statistics, etc.

The MML principle requires that the prior beliefs of the model to be fully 
specified as part of the message. 
This requirement provides a familiar link to Bayesian statistics, and allows
one to think of MML as a form of a Bayesian optimization problem that is 
deeply grounded in information theory.
Unfortunately, strict MML is generally intractable, and therefore we require
an app

@@

\bibitem[Shannon (1948)]{Shannon_1948} Shannon, C.~E.\ 1948, Bell System Technical Journal, 27, 3 % 10.1002/j.1538-7305.1948.tb01338.x

\bibitem[Schafer (1997)]{Schafer_1997} Schafer, J.~L.\ 1997, Analysis of Incomplete Multivariate Data (CRC Press) % 10.1002/(SICI)1097-0258(20000415)19:7<1006::AID-SIM384>3.0.CO;2-T

\bibitem[Wallace \& Boulton (1968)]{Wallace_1968} Wallace, C.~S. \& Boulton, D.~M.\ 1968, The Computer Journal, 11, 2 %10.1093/comjnl/11.2.185
